Design and Implementation of Probabilistic Programming Language Anglican

### 2017-02-25 paper/paper.tex
edits to section 3

A probabilistic program, or query, mostly runs deterministic code,
except for \texttt{sample} and \texttt{observe} forms, which are
probabilistic in nature. For this reason, Anglican can be mostly implemented as a regular programming language, except for the handling of these two special forms, which we refer to as checkpoints. Internally, an Anglican query is represented by a
computation in \textit{continuation passing style} (CPS)~\cite{AJ89}.
Inference algorithms exploit the CPS structure of the code to
evaluate \texttt{sample} and \texttt{observe} checkpoints in an
algorithm-specific manner\footnote{\cite{GS15} also describe a
CPS-based implementation of a probabilistic programming language.}.
The available inference algorithms include the Particle
Cascade~\cite{PWD+14}, Lightweight Metropolis-Hastings~\cite{WSG11},
Iterative Conditional Sequential Monte-Carlo (Particle Gibbs)
\cite{WVM14}, and others. Inference on Anglican queries generates a
lazy sequence of samples, which can be processed asynchronously in
Clojure for analysis, integration, and decision making.

@@

}{%
}%
Because of the checkpoints, Anglican programs must allow the
inference algorithm to step in, recording information and
affecting control flow. This can be implemented through
coroutines/cooperative multitasking, and parallel execution/preemptive
multitasking, as well as through explicit
maintenance of program continuations at checkpoints. Anglican
follows the latter option.
\iftoggle{full}{%
Clojure is a functional language, and continuation-passing style
(CPS) transformation is a well-developed technique in the area
of functional languages. Implementing a variant of CPS
transformation seemed to be the most flexible and lightweight
option --- any other form of concurrency would put a higher
burden on the underlying runtime (JVM) and the operating system.
Consequently, Anglican has been implemented as a CPS-transformed
computation with access to continuations in probabilistic
checkpoints. 
}{%
}%
The Anglican `compiler'\iftoggle{full}{, represented by a set of
functions in the \texttt{anglican.trap} namespace,}{} accepts a
Clojure subset and transforms it into a variant of CPS
representation, which allows inference algorithms to intervene
in the execution flow at probabilistic checkpoints.

Anglican is intended to co-exist with Clojure and be a part of
the source of a Clojure program. To facilitate this, Anglican
programs, or queries, are wrapped by macros\iftoggle{full}{ (defined in the
\texttt{anglican.emit} namespace)}{} which call the CPS
transformations and define Clojure values suitable for passing
as arguments to inference algorithms (\texttt{defquery},
\texttt{query}). In addition to defining entire queries,
Anglican promotes modularization of inference algorithms through
the definitions of \textit{probabilistic functions} using
\texttt{defm} (Anglican counterpart of Clojure \texttt{defn}).
Probabilistic functions are
written in Anglican, may include probabilistic forms
\texttt{sample} and \texttt{observe}, and can be seamlessly
called from inside Anglican queries, just like functions locally
defined within the same query.

>>>>

A probabilistic program, or query, mostly runs deterministic code,
except for \texttt{sample} and \texttt{observe} forms, which are
probabilistic in nature. For this reason, Anglican can be implemented
as a regular programming language, except for the handling of these
two special forms, which we refer to as checkpoints. At these
checkpoints, Anglican programs must allow the inference algorithm to
step in, recording information and affecting control flow. This can
be implemented through coroutines/cooperative multitasking, and
parallel execution/preemptive multitasking, as well as through
explicit maintenance of program continuations at checkpoints.
Anglican follows the latter option.

Internally, an Anglican query is represented by a computation in
\textit{continuation passing style} (CPS) \cite{AJ89}. The Anglican 
`compiler'\iftoggle{full}{, represented by a set of functions in the
\texttt{anglican.trap} namespace,}{} accepts a Clojure subset and 
transforms it into a variant of CPS representation, which allows 
inference algorithms to intervene in the execution flow at 
probabilistic checkpoints \footnote{\cite{GS15} also describe a
CPS-based implementation of a probabilistic programming language.}.
The available inference algorithms include the Particle Cascade
\cite{PWD+14}, Lightweight Metropolis-Hastings \cite{WSG11}, Iterative Conditional Sequential Monte-Carlo (Particle Gibbs)
\cite{WVM14}, and others. Inference on Anglican queries generates a lazy sequence of samples, which can be processed asynchronously in Clojure for analysis, integration, and decision making.

% Inference algorithms exploit the CPS structure of the code to
% evaluate \texttt{sample} and \texttt{observe} checkpoints in an algorithm-specific manner

@@

}{%
}%
% Because of the checkpoints, Anglican programs must allow the
% inference algorithm to step in, recording information and
% affecting control flow. This can be implemented through
% coroutines/cooperative multitasking, and parallel execution/preemptive
% multitasking, as well as through explicit
% maintenance of program continuations at checkpoints. Anglican
% follows the latter option.
\iftoggle{full}{%
Clojure is a functional language, and continuation-passing style
(CPS) transformation is a well-developed technique in the area
of functional languages. Implementing a variant of CPS
transformation seemed to be the most flexible and lightweight
option --- any other form of concurrency would put a higher
burden on the underlying runtime (JVM) and the operating system.
Consequently, Anglican has been implemented as a CPS-transformed
computation with access to continuations in probabilistic
checkpoints. 
}{%
}%

Anglican is intended to co-exist with Clojure and be a part of
the source of a Clojure program. To facilitate this, Anglican
programs, or queries, are wrapped by macros\iftoggle{full}{ (defined in the
\texttt{anglican.emit} namespace)}{} which call the CPS
transformations and define Clojure values suitable for passing
as arguments to inference algorithms (\texttt{defquery},
\texttt{query}). In addition to defining entire queries,
Anglican promotes modularization of probabilistic models through
the definitions of \textit{probabilistic functions} using
\texttt{defm} (Anglican counterpart of Clojure \texttt{defn}).
Probabilistic functions are
written in Anglican, may include probabilistic forms
\texttt{sample} and \texttt{observe}, and can be seamlessly
called from inside Anglican queries, just like functions locally
defined within the same query.
